Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/introduction.tex
totally fixing a typo

To support correct array programming, we propose a simple,
abstract specification language for the data access pattern of
array-loop computations. This provides a way to rule out bugs due to
indexing errors. The language design is informed by an initial
empirical study of array computations in a corpus real-world
scientific code base, totally 1.4 million lines of
code (\Cref{sec:study}). We confirm our initial hypotheses of the
ubiquity of looped array computations, but also that they have a
common form, reading from arrays in a fixed neighbourhood of
contiguous elements with a simple static pattern. From this we
designed a simple set of combinators to express common array patterns (\Cref{sec:lang}).
As an example, the Laplace program is specified in our language by:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
That is, \texttt{a} is accessed with a symmetrical pattern in its
first dimension (``centered'') to a depth of one in each direction and
its result contributes to an array
write.  The Navier-Stokes example has two specifications:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2) + centered(depth=1,dim=2)*pointed(dim=1) :: u
!= stencil forward(depth=1,dim=1)*backward(depth=1,dim=2) :: v
\end{minted}
%
The specifications requires that \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension. The specification is relatively small and abstract compared
with the code, with a small number of combinators that \emph{do not
  involve any indexing expressions}, \eg{} \texttt{a(i+1,j-1)}. This
contrasts with other specification approaches, \eg{}, deductive
verification tools such as ACSL~\citet{baudin2008acsl}, where
specifications about array computations must also be expressed in
terms of array-indexing expressions. Thus, any low-level mistakes that
could be made whilst programming complex indexing code could also be
made when writing its specification. Our specifications are
abstract to avoid the error-prone nature of index manipulation; our
specifications are more lightweight to aid its adoption in science.

>>>>

To support correct array programming, we propose a simple,
abstract specification language for the data access pattern of
array-loop computations. This provides a way to rule out bugs due to
indexing errors. The language design is informed by an initial
empirical study of array computations in a corpus of a real-world
scientific code base, totaling 1.4 million lines of
code (\Cref{sec:study}). We confirm our initial hypotheses of the
ubiquity of looped array computations, but also that they have a
common form, reading from arrays in a fixed neighbourhood of
contiguous elements with a simple static pattern. From this we
designed a simple set of combinators to express common array patterns (\Cref{sec:lang}).
As an example, the Laplace program is specified in our language by:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
That is, \texttt{a} is accessed with a symmetrical pattern in its
first dimension (``centered'') to a depth of one in each direction and
its result contributes to an array
write.  The Navier-Stokes example has two specifications:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2) + centered(depth=1,dim=2)*pointed(dim=1) :: u
!= stencil forward(depth=1,dim=1)*backward(depth=1,dim=2) :: v
\end{minted}
%
The specifications requires that \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension. The specification is relatively small and abstract compared
with the code, with a small number of combinators that \emph{do not
  involve any indexing expressions}, \eg{} \texttt{a(i+1,j-1)}. This
contrasts with other specification approaches, \eg{}, deductive
verification tools such as ACSL~\citet{baudin2008acsl}, where
specifications about array computations must also be expressed in
terms of array-indexing expressions. Thus, any low-level mistakes that
could be made whilst programming complex indexing code could also be
made when writing its specification. Our specifications are
abstract to avoid the error-prone nature of index manipulation; our
specifications are more lightweight to aid its adoption in science.
